Tropopause Evolution in a Rapidly Intensifying Tropical Cyclone: A Static Stability Budget Analysis

### 2018-01-30 paper.tex
Expanded introduction and added references.

After undergoing a remarkably rapid intensification (RI), Hurricane Patricia (2015) set a new record as the strongest tropical cyclone (TC) ever observed in the Western Hemisphere (\citeauthor{Kimberlainetal2016} \citeyear{Kimberlainetal2016}; \citeauthor{Rogersetal2017} \citeyear{Rogersetal2017}).
High-altitude dropsonde observations taken during the Tropical Cyclone Intensity (TCI) Experiment captured this RI in unprecedented detail \citep{DoyleTCI}.
These observations revealed remarkable changes in the structure of the cold-point tropopause and upper-level static stability as the storm intensified \citep{Duran+Molinari2018}.
At tropical storm intensity, shortly before RI commenced, a strong inversion layer existed just above Patricia's cold-point tropopause, which was located near 17.2 km.
During the first half of the RI period, this inversion layer weakened throughout Patricia's inner core, with the weakening most pronounced over the developing eye.
By the time the storm reached its maximum intensity, the inversion layer over the eye had disappeared almost completely, which was accompanied by an increase in the tropopause height to a level at or above the highest-available dropsonde data point (18.3 km) at two locations.
Meanwhile over the eyewall region, the static stability re-strengthened and the tropopause was limited to a level at or below 17.5 km.
The mechanisms that led to these changes in upper-level static stability and tropopause height are the subject of the current paper.

INCLUDE IN THIS INTRODUCTION DAN STERN'S POTENTIAL TEMPERATURE BUDGET PAPERS
ALSO INCLUDE KEPERT'S STATIC STABILITY BUDGET PAPER

@@

The average $N^2$ over the first day of the simulation (Fig.~\ref{fig:n2-24hr-avgs}a) indicates the presence of a weak static stability maximum just above the cold-point tropopause.
This lower-stratospheric stable layer had begun to erode during the initial spin-up period, with the maximum destabilitzation occurring at the innermost radii.
This decrease in static stability continued into the second day of the simulation (Fig.~\ref{fig:n2-24hr-avgs}b) as the storm intensified to hurricane strength (Fig.~\ref{fig:vmax+pmin}).
Destabilization was particularly pronounced over the developing eye, where the time-mean cold-point tropopause height increased by up to one km compared to the previous day.
Over the developing eyewall and rainband regions, meanwhile, the tropopause height remained nearly constant.
During the third day of the simulation (Fig.~\ref{fig:n2-24hr-avgs}c), static stability over the eye continued to decrease, and the cold-point tropopause height rose to nearly 20 km over the eye.
The tropopause sloped sharply downward outside of the 20-km radius, reaching a minimum altitude of 16.4 km only 45 km from the storm center.
This local minimum in tropopause height corresponded to the eyewall region, where upper-tropospheric static stability increased during this time period.
At larger radii, static stability began to increase in the layer immediately overlying the cold-point tropopause.
This stable layer sloped upward with radius, which corresponded to an upward-sloping tropopause radially outside of the eywall region.
Over the next 24 hours (Fig.~\ref{fig:n2-24hr-avgs}d), as the storm's maximum 10-m wind speed remained quasi-steady near 80 m s\textsuperscript{-1} (Fig.~\ref{fig:vmax+pmin}), the tropopause-layer static stability continued to increase.
%As this stable layer strengthened, its altitude rose slightly, which corresponded to a slight increase in tropopause height outside of the eyewall during this period.
%Within the upper troposphere radially outside of the eyewall, meanwhile, static stability decreased such that it was nearly neutral in a thin layer between the 120- and 150-km radii.
Within the eye, the layer between 16 and 19 km continued to destabilize, and the cold-point tropopause height increased to a level above 21.5 km.
This static stability evolution closely follows that observed in Hurricane Patricia (2015; \citeauthor{Duran+Molinari2018} \citeyear{Duran+Molinari2018}).
Since most of the static stability variability

>>>>

After undergoing a remarkably rapid intensification (RI), Hurricane Patricia (2015) set a new record as the strongest tropical cyclone (TC) ever observed in the Western Hemisphere (\citeauthor{Kimberlainetal2016} \citeyear{Kimberlainetal2016}; \citeauthor{Rogersetal2017} \citeyear{Rogersetal2017}).
High-altitude dropsonde observations taken during the Tropical Cyclone Intensity (TCI) experiment captured this RI in unprecedented detail \citep{DoyleTCI}.
These observations revealed remarkable changes in the structure of the cold-point tropopause and upper-level static stability as the storm intensified \citep{DuranMolinari2018}.

At tropical storm intensity, shortly before RI commenced, a strong inversion layer existed just above Patricia's cold-point tropopause, which was located near 17.2 km.
During the first half of the RI period, this inversion layer weakened throughout Patricia's inner core, with the weakening most pronounced over the developing eye.
By the time the storm reached its maximum intensity, the inversion layer over the eye had disappeared almost completely, which was accompanied by an increase in the tropopause height to a level at or above the highest-available dropsonde data point (18.3 km) at two locations.
Meanwhile over the eyewall region, the static stability re-strengthened and the tropopause was limited to a level at or below 17.5 km.
The mechanisms that led to these changes in upper-level static stability and tropopause height are the subject of the current paper.

Despite the importance of tropopause-layer thermodynamics in theoretical models of hurricanes (\citeauthor{EmanuelRotunno2011} \citeyear{EmanuelRotunno2011}; \citeauthor{Emanuel2012} \citeyear{Emanuel2012}), few papers have examined the upper-tropospheric evolution of TCs.
\cite{KomaromiDoyle2017} found that stronger TCs tended to have a higher and warmer tropopause over their inner core than weaker TCs.
Their results are consistent with the evolution observed over the inner core of Hurricane Patricia, in which the tropopause height increased and the tropopause temperature warmed throughout RI \citep{DuranMolinari2018}.
The simulations of \cite{OhnoSatoh2015} suggested that the development of an upper-level warm core within the eye acted to decrease the static stability near the tropopause.
Although the mechanisms that drive this static stability evolution have not been examined explicitly, the potential temperature ($\theta$) budget analysis of \cite{SternZhang2013} examined the development of the TC warm core.
They found that radial and vertical advection both play important roles in warm core development throughout RI, with subgrid-scale diffusion becoming particularly important during the later stage of RI.

The analysis herein is based upon that of \cite{SternZhang2013}, except using a static stability budget similar to that of \cite{Kepertetal2016} rather than a $\theta$ budget.

@@

The average $N^2$ over the first day of the simulation (Fig.~\ref{fig:n2-24hr-avgs}a) indicates the presence of a weak static stability maximum just above the cold-point tropopause.
This lower-stratospheric stable layer had begun to erode during the initial spin-up period, with the maximum destabilitzation occurring at the innermost radii.
This decrease in static stability continued into the second day of the simulation (Fig.~\ref{fig:n2-24hr-avgs}b) as the storm intensified to hurricane strength (Fig.~\ref{fig:vmax+pmin}).
Destabilization was particularly pronounced over the developing eye, where the time-mean cold-point tropopause height increased by up to one km compared to the previous day.
Over the developing eyewall and rainband regions, meanwhile, the tropopause height remained nearly constant.
During the third day of the simulation (Fig.~\ref{fig:n2-24hr-avgs}c), static stability over the eye continued to decrease, and the cold-point tropopause height rose to nearly 20 km over the eye.
The tropopause sloped sharply downward outside of the 20-km radius, reaching a minimum altitude of 16.4 km only 45 km from the storm center.
This local minimum in tropopause height corresponded to the eyewall region, where upper-tropospheric static stability increased during this time period.
At larger radii, static stability began to increase in the layer immediately overlying the cold-point tropopause.
This stable layer sloped upward with radius, which corresponded to an upward-sloping tropopause radially outside of the eywall region.
Over the next 24 hours (Fig.~\ref{fig:n2-24hr-avgs}d), as the storm's maximum 10-m wind speed remained quasi-steady near 80 m s\textsuperscript{-1} (Fig.~\ref{fig:vmax+pmin}), the tropopause-layer static stability continued to increase.
%As this stable layer strengthened, its altitude rose slightly, which corresponded to a slight increase in tropopause height outside of the eyewall during this period.
%Within the upper troposphere radially outside of the eyewall, meanwhile, static stability decreased such that it was nearly neutral in a thin layer between the 120- and 150-km radii.
Within the eye, the layer between 16 and 19 km continued to destabilize, and the cold-point tropopause height increased to a level above 21.5 km.
This static stability evolution closely follows that observed in Hurricane Patricia (2015; \citeauthor{DuranMolinari2018} \citeyear{DuranMolinari2018}).
Since most of the static stability variability
